Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-09-28 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
complete 4 and revising Ch5

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline {Burner} & \multicolumn{2}{c}{Natural Gas} & \multicolumn{2}{c}{Gasoline} & \multicolumn{2}{c|}{PU Foam} \\
		\hline  Position & Temp~(C) & HF~(kW/m$^{2}$) & Temp~(C) & HF~(kW/m$^{2}$) & Temp~(C) & HF~(kW/m$^{2}$) 	\\
		\hline \hline 0.61~m & 95 & 5 & 106 & 7 & 71 & 6 \\
		\hline 0.31~m & 164 & 14 & 232 & 24 & 148 & 16	\\
		\hline 0.15~m & 257 & 33 & 356 & 45 & 266 & 39	\\
		\hline 0.00~m & 487 & 43 & 621 & 70 & 526 & 65 	\\
		\hline
	\end{tabular}
	\caption[Summary of instrumented non-combustible wall experiments]{Summary of peak temperatures and heat fluxes from the instrumented non-combustible wall experiments.  The estimated combined expanded uncertainty for temperature in these experiments was $\pm$15~\% and $\pm$8~\% for the heat flux.}
	\label{tab:TW_Summary}
\end{table}

@@

A summary of the peak average temperatures and heat fluxes at the wall is provided in Table~\ref{tab:IWGB_Summary}.

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline {Burner} & \multicolumn{2}{c}{Natural Gas} & \multicolumn{2}{c}{Gasoline} & \multicolumn{2}{c|}{PU Foam} \\
		\hline  Position & Temp~(C) & HF~(kW/m$^{2}$) & Temp~(C) & HF~(kW/m$^{2}$) & Temp~(C) & HF~(kW/m$^{2}$) 	\\
		\hline \hline 0.61~m & 132 & 4 & 109 & 6 & 76 & 4 \\
		\hline 0.31~m & 160 & 11 & 185 & 18 & 151 & 12	\\
		\hline 0.15~m & 250 & 23 & 377 & 57 & 244 & 25	\\
		\hline 0.00~m & 539 & 55 & 590 & 61 & 415 & 59 	\\
		\hline
	\end{tabular}
	\caption[Summary of instrumented gypsum wallboard experiments]{Summary of peak temperatures and heat fluxes from the instrumented gypsum wallboard experiments}
	\label{tab:IWGB_Summary}
\end{table}

Figure~\ref{IWGB_NG_patterns} provides a comparison between the area of discoloration of the paint with the natural gas burner positioned at 0.15~m away from the wall versus the area of consumed and penetrated paint/paper layer which exposed the gypsum underneath.

@@

@@

@@

\begin{figure}[p]
	\centering
	\includegraphics[trim=5.0in 1.7in 5.0in 15.0in, clip=true, width=0.4\columnwidth]{../Figures/IWGB_PUF8_0m_pattern2}
	\includegraphics[trim=5.0in 3.5in 5.0in 13.0in, clip=true, width=0.4\columnwidth]{../Figures/IWGB_PUF9_0m_pattern} \\
	\caption[Photographs of the fire effects on the instrumented gypsum wallboard exposed to replicate polyurethane foam fires]{Photographs of the fire effects on the instrumented gypsum wallboard exposed to replicate polyurethane foam fires with the burner positioned 0.0~m from the wall.}
	\label{IWGB_PUF_0m_patterns}
\end{figure}

@@

The purpose of this chapter is to develop and compare data from bench scale apparatus that could be used to analyze fire damage to gypsum wallboard.
This type of testing might be practical for a fire investigator to use based on the low cost, relative to a full-scale fire test, and the small size of the samples needed for the tests.  In the previous sections of this study, burners were used to examine the repeatability of each fire's characteristics, the amount of heat transfered to a wall as a function of distance and the damage to a section of painted gypsum wallboard.  While the data generated from those experiments provide valuable insight, it would be of value to be able to develop assessments of fire patterns based on data from one or a series of bench scale tests.

@@

For the ignition tests, an electric spark igniter was positioned over the center of the horizontal samples to serve as a pilot.  Combustion products and dilution air were extracted through the hood and exhaust duct by a high temperature fan. The flow rate can be adjusted between 0.01 and 0.03~m$^3$/s. The volumetric flow rate was kept constant during testing. The sample was mounted on a load cell to determine mass loss rate during a test, and smoke obscuration was measured using a laser light source. The gas flow rate in the exhaust duct was calculated from the pressure drop across and temperature at an orifice plate in the duct. Finally, the concentrations of oxygen, carbon dioxide, carbon monoxide, were measured using appropriate instruments. Heat release rate was calculated from the gas concentration and mass flow measurements.

@@

Prior to conducting experiments with the gypsum wallboard samples, heat flux gauge measurements taken at the center position of the sample were used to determine the positions on the trolley track for a specific radiant exposure. The positions were marked for 10, 12.5, 15, 17.5, 20, 22.5, and 25~kW/m$^2$. The area surrounding the center of the painted gypsum wallboard sample maintained the desired heat flux within at least 70~mm on either side of the center point and 100~mm above and below the center point.  In other words, the area around the center point of the sample had heat flux measures similar, within the uncertainty of the device, to the center point measurement. Outside of the central area of the samples, the heat flux would decrease as the position approached the edge of the sample.  The center point heat flux measures vs the distance from the radiating surface is provided in Figure~\ref{Rad_Panel_Cal}

>>>>

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline {Burner} & \multicolumn{2}{c}{Natural Gas} & \multicolumn{2}{c}{Gasoline} & \multicolumn{2}{c|}{PU Foam} \\
		\hline  Position & Temp~($^\circ$C) & HF~(kW/m$^{2}$) & Temp~($^\circ$C) & HF~(kW/m$^{2}$) & Temp~($^\circ$C) & HF~(kW/m$^{2}$) 	\\
		\hline \hline 0.61~m & 95 & 5 & 106 & 7 & 71 & 6 \\
		\hline 0.31~m & 164 & 14 & 232 & 24 & 148 & 16	\\
		\hline 0.15~m & 257 & 33 & 356 & 45 & 266 & 39	\\
		\hline 0.00~m & 487 & 43 & 621 & 70 & 526 & 65 	\\
		\hline
	\end{tabular}
	\caption[Summary of instrumented non-combustible wall experiments]{Summary of averaged peak temperatures and averaged heat fluxes from the instrumented non-combustible wall experiments.  The estimated combined expanded uncertainty for temperature in these experiments was $\pm$15~\% and $\pm$8~\% for the heat flux.}
	\label{tab:TW_Summary}
\end{table}

@@

A summary of the peak average temperatures and heat fluxes at the wall is provided in Table~\ref{tab:IWGB_Summary}. While the values in this summary table differ from the values in Table~\ref{tab:TW_Summary} a closer look reveals that most of the values are not ``significantly '' different.  In other words the range of expanded uncertainty values overlap.  For example at 0.0~m position with natural gas, the average peak temperature was 487~$^\circ$C~$pm$73~$^\circ$C (15~\%). So the upper bound would be 560~$^\circ$C, which includes the corresponding averaged peak value from the instrumented gypsum wallboard experiments of 539~$^\circ$C.  While all of the values do not have overlapping uncertainties, the point is that none of the averaged peak values are ``that '' different.

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline {Burner} & \multicolumn{2}{c}{Natural Gas} & \multicolumn{2}{c}{Gasoline} & \multicolumn{2}{c|}{PU Foam} \\
		\hline  Position & Temp~($^\circ$C) & HF~(kW/m$^{2}$) & Temp~($^\circ$C) & HF~(kW/m$^{2}$) & Temp~($^\circ$C) & HF~(kW/m$^{2}$) 	\\
		\hline \hline 0.61~m & 132 & 4 & 109 & 6 & 76 & 4 \\
		\hline 0.31~m & 160 & 11 & 185 & 18 & 151 & 12	\\
		\hline 0.15~m & 250 & 23 & 377 & 57 & 244 & 25	\\
		\hline 0.00~m & 539 & 55 & 590 & 61 & 415 & 59 	\\
		\hline
	\end{tabular}
	\caption[Summary of instrumented gypsum wallboard experiments]{Summary of averaged peak temperatures and heat fluxes from the instrumented gypsum wallboard experiments.  The estimated combined expanded uncertainty for temperature in these experiments was $\pm$15~\% and $\pm$8~\% for the heat flux.} 
	\label{tab:IWGB_Summary}
\end{table}

Figure~\ref{IWGB_NG_patterns} provides a comparison between the area of discoloration of the paint with the natural gas burner positioned at 0.15~m away from the wall versus the area of consumed and penetrated paint/paper layer which exposed the gypsum underneath.  The peak height of the consumed and penetrated paper, i.e. the burned region, was 0.8~above the burner.  This is consistent with the prediction based on the lower heat flux ignition value found in the literature and the average peak heat flux from the non-combustible was experiments.

@@

Figures~\ref{IWGB_Gas_patterns} and~\ref{IWGB_Gas_0m_patterns} show the progression of thermal damage caused by gasoline fueled burner at 0.15~m, 0.08~m, and 0.0~m away from the gypsum wallboard.  When the burner was positioned at 0.15~m away from the wall, a semi-circular narrow burn line formed with the lowest point approximately 0.1~m above the burner and extending up to the 0.4~m above the burner on the left side.  The formation of the burned area was captured on film and is shown in Figure~\ref{IWGBGAS6_seq}.  The burned region on the gypsum wallboard from exposure to the gasoling fire position 0.08~m away began at 0.2~m above the burner and extended up to 0.6~m. The width of the burned region was approximately one half of the burner width at it's maximum.  Figure~\ref{IWGB_Gas_0m_patterns} shows the fire pattern formed by gas burner when it is tight against the gypsum wallboard.  Two views are shown.  The image on the left shows the wall shortly after the fire went out.  The burned area came be discerned, but a close look is required.  In the photograph on the right, the same piece of gypsum wallboard was lightly brushed off, which revealed a clearer view of the area where the paint and paper were burned off.  The peak height of that fire pattern is just short of 0.8~m. The maxim width is approximately the full width of the burner.

@@

Figures~\ref{IWGB_PUF_patterns} and~\ref{IWGB_PUF_0m_patterns} show the progression of thermal damage caused by the polyurethane fueled burner at 0.15~m, 0.08~m, and 0.0~m away from the instrumented gypsum wallboard.  The burner positioned at 0.15~m from the wall did not transfer enough energy to cause any thermal damage, the only damage to the wall is from soot deposition.  When the burner was located 0.08~m away from the wall, the paint/paper finish over the gypsum was discolored and heavily sooted but the was no consumption or penetration of the paper finish layer.  Three replicate experiments were conducted with the polyurethane foam fueled burner tight against the instrumented gypsum wallboard. The resulting patterns are shown in Figure~\ref{IWGB_PUF_0m_patterns}.  The average height of the burned region of the painted paper finish on the wallboard was 0.3~m above the top of the burner. The maximum width of the burned region was approximately equal to the width of the burner.

@@

\begin{figure}[p]
	\centering
	\includegraphics[trim=0.0in 0.0in 0.0in 0.0in, clip=true, width=0.3\columnwidth]{../Figures/IWGB_PUF7_0m_crop}
	\includegraphics[trim=5.0in 1.7in 5.0in 15.0in, clip=true, width=0.3\columnwidth]{../Figures/IWGB_PUF8_0m_pattern2}
	\includegraphics[trim=5.0in 3.5in 5.0in 13.0in, clip=true, width=0.3\columnwidth]{../Figures/IWGB_PUF9_0m_pattern} \\
	\caption[Photographs of the fire effects on the instrumented gypsum wallboard exposed to replicate polyurethane foam fires]{Photographs of the fire effects on three instrumented gypsum wallboards exposed to replicate polyurethane foam fires with the burner positioned 0.0~m from the wall.}
	\label{IWGB_PUF_0m_patterns}
\end{figure}

With this set of experiments, fire patterns as the result of thermal damage have finally generated in a few cases.  To recap our crude predictions from the previous chapter, the natural gas burner had the potential to cause a fire pattern with a height between 0.6~m and 0.8~m.  With the natural gas burner against the wall, a pattern with a maximum burn area of 0.8~m in height was generated.  From Chapter III, the gasoline had a potential to create a fire pattern with a maximum height between 0.8~m and 1.0~m.  In the single instrumented gypsum wallboard experiment with the gasoline burner located against the wall, the maximum height of the burned region was just short of 0.8~m.  The prediction for the potential height of the fire pattern that could be created with the polyurethane burner against the wall was between 0.4~ and 0.6~m.  With the polyurethane burner three replicate experiments were conducted.  The range of the maximum heights of the burned region was 0.2~m to 0.35~m.  The average peak burned region height was 0.3~m. The uncertainty of the repeatibility of the burn patterns has not been assessed yet, but it appears that the crude predictions from Ch III provided a ``ballpark'' estimate of the maximum height of the burned region.

@@

The purpose of this chapter is to develop and compare data from bench scale apparatus that could be used to analyze fire damage to gypsum wallboard.
This type of testing might be practical for a fire investigator to use based on the low cost, relative to a full-scale fire test, and the small size of the samples needed for the tests.  In the previous sections of this study, burners were used to examine the repeatability of each fire's characteristics, the amount of heat transfered to a wall as a function of distance and the damage to a section of painted gypsum wallboard.  While the data generated from those full scale laboratory experiments provide valuable insight, it would be better to be able to develop assessments of fire patterns based on data from one or a series of bench scale tests in an effort to validate or improve the existing data in the literature.

@@

For the ignition tests, an electric spark igniter is positioned over the center of the horizontal samples to serve as a pilot.  Combustion products and dilution air were extracted through the hood and exhaust duct by a high temperature fan. The flow rate can be adjusted between 0.01 and 0.03~m$^3$/s. The volumetric flow rate was kept constant during testing. The sample was mounted on a load cell to determine mass loss rate during a test, and smoke obscuration was measured using a laser light source. The gas flow rate in the exhaust duct was calculated from the pressure drop across and temperature at an orifice plate in the duct. Finally, the concentrations of oxygen, carbon dioxide, carbon monoxide, were measured using appropriate instruments. Heat release rate was calculated from the gas concentration and mass flow measurements.

@@

Prior to conducting experiments with the gypsum wallboard samples, heat flux gauge measurements taken at the center position of the sample were used to determine the positions on the trolley track for a specific radiant exposure. The positions were marked for 10, 12.5, 15, 17.5, 20, 22.5, and 25~kW/m$^2$. The area surrounding the center of the painted gypsum wallboard sample maintained the desired heat flux within at least 70~mm on either side of the center point and 100~mm above and below the center point.  In other words, the area around the center point of the sample had heat flux measures similar, in other words within the uncertainty of the heat flux sensor, to the center point measurement. Outside of the central area of the samples, the heat flux would decrease as the position approached the edge of the sample.  The center point heat flux measures vs the distance from the radiating surface is provided in Figure~\ref{Rad_Panel_Cal}
